A Mechanised Library for Generic (Co)Programming

### 2024-03-14 paper/paper.tex
Introduction

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming and recursion schemes provide powerful abstractions for
reasoning about program equivalences and deriving optimisations. However,
formalizing such abstractions in Coq often requires compromises, such as
imposing performance penalties, requiring the assumption of additional axioms,
or introducing unsafe casts into extracted code (for example, \ocaml{Obj.magic}
in OCaml). This paper presents a fully axiom-free formalization of hylomorphisms
(a recursion scheme capturing divide-and-conquer algorithms) in Coq, along with
their algebraic laws. The formalization allows for the extraction of clean,
natural-looking OCaml code for a variety of recursive programs.
% while enabling
% formal reasoning about program equivalences using the laws of hylomorphisms.

The key contribution of this paper is an axiom-free framework for generic
programming with recursion schemes in Coq, focusing on proving algebraic laws,
reasoning about program equivalences, and extracting safe, idiomatic code. With
this framework, we formalize a series of divide-and conquer, dynamic
programming, and mutual recursion programs as hylomorphisms. We demonstrate that
the extracted OCaml code for the programs formalized in our framework is
efficient, resembles code that a human programmer might write, and contains no
uses of \ocaml{Obj.magic}. We also present a machine-checked proof of the
well-known short-cut fusion optimisation.

By avoiding undafe coersions in the extracted code, this work provides a solid
foundation for developing efficient, verified recursive programs with strong
guarantees about the safety and interoperability of the generated OCaml code.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity.  For recursion, this is done by trying to (automatically) infer
which argument in the recursive call gets smaller with respects to the original
input argument.  For productivity, the algorithm checks that the (co)recursive
call appears directly under a constructor to make sure that this function
always produce at least one element after each recursive step.

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorith, written in OCaml below:
\begin{minted}{OCaml}
  let divide = function
    | [] -> None
    | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
  let rec qsort xs = match divide xs with
    | None -> []
    | Some (x, (smaller, larger)) -> qsort smaller @ (x::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

The main approach for implementing nonstructural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are coupled with
\emph{termination proofs}. Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}. A benefit of
this approach is that, extracting verified nonstructural recursive functions to
OCaml will erase the termination proofs, and so produce code that will be
closer to what a programmer may have written directly in OCaml. However,
reasoning about program equivalences requires dealing with such termination
proofs, and it is common practice to use custom reduction lemmas that will be
used extensively when proving properties  about the nonstructural recursive
functions.

A better approach for reasoning about program equivalences comes from
\emph{structured recursion schemes}, and their associated algebraic laws.  In
particular, \ocaml{qsort} is an instance of a \emph{divide-and-conquer}
algorithm which, in the functional programming literature is known as a
hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
that every recursive program can be turned into an hylomorphism, which implies
that, ``\emph{every recursion scheme can be formalised as an hylomorphism}''.
Furthermore, the laws that are satisfied by hylomorphisms ease reasoning about
program equivalences, since they can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to (semi-)automatic
parallelisations~\cite{Gibbons96:Third, farmsCastro}.

Despite the benefits of hylomorphisms for reasoning about program equivalences,
not many approaches mechanise structured recursion schemes in proof assistants.
A recent approach by Abreu et al.~\cite{AbreuDHJMS23} encode an algebraic
approach to divide-and-conquer computations, in which termination is entirely
enforced by the typing discipline of the recursion scheme. However, although
their approach solves the problem of termination proofs, as well as the
performance of the code that is run \emph{within Coq}, it no longer allows the
extraction of clean, naturally-looking code.  By naturally-looking and clean,
we mean code that (1) preserves the recursive structure intended by the
programmer; and (2) does not lead to unsafe casts \ocaml{Obj.magic} in the
generated code.

In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function
  | [] -> []
  | h :: t ->
    let (l, r) = partition (fun x0 -> (<=) x0 h) t in
    let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
    app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}

The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a strong focus on proving their algebraic laws and 
    program equivalences, and clean code extraction.
  \item We formalise algorithms for divide and conquer, dynamic programming and
    mutual recursion.
  \item We verified the well-known short-cut fusion optimisation in Coq and
    extract the resulting code to OCaml.
\end{itemize}

\dcas{Rewritten up to here}
In
\ocaml{quicksort} above, a pen-and-paper technique to prove termination would
be to show that \ocaml{divide} is a \emph{well-founded coalgebra} on the
functor $(1 + A \times - \times -)$ for a fixed type $A$. Here the term
``well-founded'' refers to the fact that the infinite iteration of this
coalgebra will produce a \emph{finite} binary tree with nodes in $A$.

More generally, this is an instance of a \emph{divide-and-conquer} algorithm.
Such algorithms are structured in three parts: first, the input is broken down
in smaller sub-problems by means of a well-founded coalgebra. Secondly, these
sub-problems are computed recursively and, finally, the results of the recursive
call are put together by a an algebra. This particular way of doing recursion
has been named in the functional programming literature as a
hylomorphism~\cite{MeijerFP91, HuIT96}.

Many years later Hinze et al.~\cite{HinzeWG15} showed that every recursive
program can be turned into an hylomorphism by means of the conjugate rule. The
essence of this result is that every complex recursion scheme can be explained
by its associated \emph{adjunction} which establishes a way of reducing the
proof obligation for the complex recursion scheme into the proof obligation for
a basic hylomorphism. This can be paraphrased by saying
\begin{center}
  ``\emph{every recursion scheme can be formalised as an hylomorphism}''
\end{center}

This result has given a much needed unifying view of recursion scheme which
helped understanding their nature and also in providing a unified way of
reasoning about recursive programs by means of fusion laws.

\dcas{My take would be different here from now on. See below:}

Recursion schemes enjoy generic algebraic laws for program manipulation. Many
common optimisations and transformations are instances of such laws. Examples of
this include shortcut deforestation~\cite{TakanoM95}, and
parallelisations~\cite{Gibbons96:Third}. \dcas{TODO: add more examples?}

With this work, we study the formalisation of hylomorphism laws in proof
assistants like Coq to derive efficient and correct recursive algorithms by
program calculation techniques. Furthermore, our encoding of hylomorphisms
allows Coq's code extraction mechanism to produce OCaml code that resembles what
a programmer would have writen manually.

\dcas{END (TODO: merge with Marco's. I would omit (or take parts of the) below)}

There is a tension:
\begin{itemize}
  \item On one hand we can allow general recursion and the employ recursion
schemes to help the programmer structuring programs so that they terminate
  \item on the other hand we could forbid non-termination, but at this point
there is no algorithm to tell termination and non-termination apart.
\end{itemize}

The type theoretical approach instead is to reject all definitions which cannot
be proven well-defined. It is safer, but incomplete.

With this work we intend to begin a study of the applications of recursion
schemes that goes beyond functional programming, that is to dependent type
theories such as a the ones employed in proof assistants like Agda or Coq where
non-termination is outright rejected.

This has been a long-standing challenge to come up with a sound algorithm which
rejects all programs that do not terminate (soundness) and accepts as many
well-defined programs as possible (completeness). Our aim is to provide an
additional tool to the every-day programmer to be able to extend the range of
recursive programs we can write.
\mvol{What is the ``every-day programmer'' here? Are they working in Coq/Agda?}

\dcas{Note: the below is useful to keep in the intro}
One of the challenges here is to be able to keep the theory of recursions
schemes as general as possible while maintaining its practical usefulness. In
order to do this, we have to make some design choices which differ a bit from
the general theory.  For instance, formalising the conjugate rule would require
us to mechanise a non-trivial fragment of category theory inside the prover's
logic.  Secondly, the general conjugate rule is a statement of correspondence of
arrows rather than a proof that these programs actually exists. In functional
programming languages like for instance Haskell this is not a problem, since
every functor is strong and has a fixed-point.  On the other hand, in a type
theory we have to restrict to those so-called polynomial functors. In this work
we do not make use of these assumptions. By restricting ourselves to polynomial
functors given in terms of containers~\cite{AbbottAG05} we are able to prove the
recursion schemes within the proof assistant and at the same time to maintain a
certain level of data abstraction.

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.
\mpav{Moved this from Section 3}

\mvol{Need some transiiton here to build up to listing the goals and contributions of the paper. Also, ``axiom-free'' and ``obj.magic'' come out of nowhere here, so both need to be established earlier in the introduction.}
The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \ocaml{Obj.magic}.
\end{enumerate}
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}

The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
schemes in Coq
  \item We formalise algorithms for divide and conquer, dynamic programming and
mutual recursion
  \item We validated optimisations in Coq and extract the resulting code to
OCaml.
\end{itemize}

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction
  for reasoning about program equivalences and deriving optimisations which has
  been successfully applied to functional programming. However, formalising
  this technique in a type theory, (e.g. Coq) often requires compromises, such
  as imposing performance penalties, requiring the assumption of additional
  axioms, or introducing unsafe casts into extracted code (for example,
  \ocaml{Obj.magic} in OCaml).

  This paper presents a formalisation of hylomorphisms (a recursion scheme
  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
  laws. The key contribution of this paper is that this formalisation is fully
  axiom-free, and it allows for the extraction of safe, clean, idiomatic OCaml
  code for a variety of recursive programs. We formalise a series of divide-and
  conquer, dynamic programming, and mutual recursion programs and demonstrate
  that the extracted OCaml code for the programs formalised in our framework is
  efficient, resembles code that a human programmer would write, and contains
  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
  of the well-known short-cut fusion optimisation.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity.  For recursion, this is done by automatically inferring
which argument in the recursive call gets smaller with respects to the original
input argument.  For productivity, the algorithm checks that the corecursive
call appears directly under a constructor to make sure that this function
always produces at least one element after each recursive step.
This implies that some functions, though well-defined, cannot be accepted by
the proof assistant. Examples of this include common sorting algorithms, such
as Quicksort and Mergesort.
%One example is Quicksort, written in OCaml
%below:
\begin{minted}{OCaml}
let rec qsort xs = match divide xs with | None -> []
 | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

The main approach for implementing nonstructural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are coupled with
\emph{termination proofs}.  Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
one approach is using the \coq{Fix} combinator, in which recursion is done on
an \emph{accesibility} predicate on the input, which is a proof that there are
no infinitely decreasing chains.}. A benefit of this approach is that,
extracting verified nonstructural recursive functions to OCaml will erase the
termination proofs, and so produce code that will be closer to what a
programmer may have written directly in OCaml. However, reasoning about program
equivalences requires dealing with such termination proofs, and it is common
practice to use custom reduction lemmas that will be used extensively when
proving properties about nonstructural recursive functions.

\emph{Structured recursion schemes} and their associated algebraic laws provide
an easier framework for reasoning about program equivalences.  In particular,
\ocaml{qsort} is an instance of a \emph{divide-and-conquer}
algorithm which, in the functional programming literature is known as a
hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
that every recursive program (recursion scheme) can be formalised as a hylomorphism.
Furthermore, hylomorphism laws can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to semi-automatic
parallelisations~\cite{Gibbons96:Third, farmsCastro}.

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of clean, naturally-looking code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}
